Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-06 paper/paper.tex
more details on probabilistic forms

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Semantics of probabilistic forms in Anglican which facilitates full
        separation of probabilistic programs from inference algorithms.
    \item Stateless representation of random processes, seamlessly integrated
        into pure functional computation.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
\end{itemize}

@@

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textbf{checkpoints} --- Clojure values of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The types contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called:
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
expressions are constructors for Clojure records, and they take 
values of their fields as arguments.

@@

\iftoggle{full}{%
In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.
}{%
}%

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a Clojure keyword (a symbolic constant starting with colon)
related to the algorithm name, such as \texttt{:lmh} for Lightweight
Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
%If needed, the query
%can be defined anonymously, right in the argument list of a call
%to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query x) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
%The value is destructured via pattern match in case of a structured binding.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (sample (normal mean sd)))

(def samples (infer :lmh my-query [1.0 3.0]))
\end{lstlisting}
Finally, any number of auxiliary arguments can be
passed to \texttt{infer}.
%Finally, an indefinite number of auxiliary arguments can be
%passed to \texttt{infer}.
By convention, the arguments should be
keyword arguments, and are interpreted in the algorithm-specific
manner.

\subsection{Internals of an Inference Algorithm}
\label{seq:internals}

>>>>

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Specification of probabilistic forms in Anglican which facilitates 
        separation of probabilistic programs from inference algorithms.
    \item Stateless representation of random processes, seamlessly integrated
        into pure functional computation.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
\end{itemize}

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved through returning \textbf{checkpoints} --- Clojure
values of the corresponding types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
Checkpoints expose the program state to the inference algorithm,
and the updated state is re-injected into the computation when
the continuation is called:
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the CPS-transformed
expressions are constructors for Clojure records, and they take 
values of their fields as arguments. How the state is updated, 
the number of times the continuation is called, and the value
passed to the continuation of \texttt{sample} depend on the
inference algorithm executing the program.
Section~\ref{sec:inference} provides more detail on internals of
inference algorithms and their interaction with probabilistic
programs.

\iftoggle{full}{%
\smallskip
\begin{center}
*\qquad*\qquad*
\end{center}

In addition to the checkpoints, there are a few other special
forms ---  \texttt{store}, \texttt{retrieve}, \texttt{mem} ---
which modify program state. These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.
}{%
}%

@@

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a Clojure keyword (a symbolic constant starting with colon)
related to the algorithm name, such as \texttt{:lmh} for Lightweight
Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query x) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (sample (normal mean sd)))

(def samples (infer :lmh my-query [1.0 3.0]))
\end{lstlisting}
Finally, any number of auxiliary arguments can be
passed to \texttt{infer}.
By convention, the arguments should be
keyword arguments, and are interpreted in the algorithm-specific
manner.

\subsection{Internals of an Inference Algorithm}
\label{sec:internals}
